---
title: "Categorical and mixed-indicator models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Categorical and mixed-indicator models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

## Scope

This guide covers ordered and mixed indicators. The implementation covers
lavaan's DWLS and ULS families, one or several groups, and listwise or pairwise
missingness, and is validated against an independent implementation.

The correction uses the unscaled estimator statistic and the biased spectrum
that lavaan exposes through `lavInspect(fit, "UGamma")`. The Du--Bentler `UG`
gamma and the continuous-data `RLS` statistic do not apply.

```{r setup}
library("semTests")
library("lavaan")

model <- "
  visual  =~ x1 + x2 + x3
  textual =~ x4 + x5 + x6
  speed   =~ x7 + x8 + x9
"

ordered_names <- paste0("x", 1:9)
ordinal_data <- HolzingerSwineford1939
ordinal_data[ordered_names] <- lapply(
  ordinal_data[ordered_names],
  function(x) ordered(cut(
    x,
    breaks = c(-Inf, quantile(x, c(.33, .67)), Inf),
    include.lowest = TRUE
  ))
)
```

## Overall goodness of fit

lavaan's `WLSMV` shortcut fits the DWLS family and requests a
mean-and-variance-adjusted test:

```{r ordinal-fit}
fit_ordinal <- cfa(
  model, ordinal_data,
  ordered = ordered_names,
  estimator = "WLSMV"
)

pvalues(fit_ordinal, c("STD", "SB", "SS", "ALL", "PEBA4"))
```

The output provenance records both the base estimator (`DWLS`) and the
requested estimator (`WLSMV`), together with the categorical parameterization.

```{r ordinal-provenance}
attr(pvalues(fit_ordinal, "PEBA4"), "semtests")
```

For `SB` and `SS`, the result can be sanity-checked against lavaan's public
robust corrections. The other methods use more of the spectrum than the trace
and squared trace used by these classical corrections.

## Estimator families

The accepted shortcuts include:

* DWLS, WLSM, WLSMV, and WLSMVS
* ULS, ULSM, and ULSMV

Full WLS/ADF is *refused*. Its weight is already the inverse asymptotic moment
covariance, so the spectrum is the identity and every correction would reduce to
the ordinary chi-square -- the robust p-values add nothing, so semTests declines
the fit rather than return a naive test in robust clothing.

## Mixed indicators

Only variables named in `ordered` are treated as categorical. The remaining
indicators stay continuous:

```{r mixed-fit}
mixed_names <- paste0("x", 1:6)
mixed_data <- HolzingerSwineford1939
mixed_data[mixed_names] <- lapply(
  mixed_data[mixed_names],
  function(x) ordered(cut(
    x,
    breaks = c(-Inf, quantile(x, c(.33, .67)), Inf),
    include.lowest = TRUE
  ))
)

fit_mixed <- cfa(
  model, mixed_data,
  ordered = mixed_names,
  estimator = "WLSMV"
)
pvalues(fit_mixed, "PEBA4")
```

The threshold, covariance, and mean components of mixed sample statistics make
the moment space larger than in a continuous covariance-only analysis. Using
lavaan's inspected spectrum avoids reconstructing these conventions inside
`semTests`.

## Pairwise missingness

With pairwise categorical estimation, lavaan estimates each sample statistic
from the observations available for that pair:

```{r pairwise-fit}
pairwise_data <- ordinal_data
set.seed(314)
for (variable in ordered_names) {
  pairwise_data[sample(nrow(pairwise_data), 35), variable] <- NA
}

fit_pairwise <- cfa(
  model, pairwise_data,
  ordered = ordered_names,
  estimator = "WLSMV",
  missing = "pairwise"
)
pvalues(fit_pairwise, c("SB", "SS", "PEBA4"))
```

`semTests` follows lavaan's pairwise sample-statistic and `UGamma`
calculation. The validity of pairwise deletion depends on the missingness
process, so its assumptions deserve a separate justification. Pairwise
estimation does not provide FIML inference.

## Nested categorical models

Categorical comparison uses Satorra's 2000 restriction-map construction with
the delta method:

```{r categorical-nested}
constrained <- "
  visual  =~ x1 + a*x2 + a*x3
  textual =~ x4 + b*x5 + b*x6
  speed   =~ x7 + x8 + x9
"

m1 <- cfa(
  model, ordinal_data,
  ordered = ordered_names, estimator = "WLSMV"
)
m0 <- cfa(
  constrained, ordinal_data,
  ordered = ordered_names, estimator = "WLSMV"
)

pvalues_nested(
  m0, m1,
  method = "2000", A.method = "delta",
  tests = c("SB", "SS", "PALL")
)
```

Both fits must use the same requested estimator, parameterization, variables,
groups, raw data, and missingness mask. Method 2001 and the exact restriction
map are intentionally rejected for this family.

## Practical checks

Before interpreting the result, give the fit a quick check:

1. Verify convergence and adequate category counts.
2. Inspect sparse thresholds and pairwise observation counts.
3. Confirm that the two nested fits use identical data and options.
4. Compare `SB` and `SS` with lavaan as a trace-level sanity check.
5. Report pairwise missingness and the categorical estimator choice as
   analysis assumptions.

## References

Satorra, A. (2000). Scaled and adjusted restricted tests in multi-sample
analysis of moment structures. In R. D. H. Heijmans, D. S. G. Pollock, &
A. Satorra (Eds.), *Innovations in Multivariate Statistical Analysis*
(pp. 233–247). Kluwer Academic.
<https://doi.org/10.1007/978-1-4615-4603-0_17>
