---
title: "Categorical and mixed-indicator models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Categorical and mixed-indicator models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

## Scope

This guide covers ordered and mixed indicators. The implementation covers
lavaan's DWLS and ULS families, one or several groups, and listwise or pairwise
missingness. The supported paths have also been checked against a separate
implementation.

For these models, `semTests` uses lavaan's unscaled statistic and the
eigenvalues stored in `lavInspect(fit, "UGamma")`. The Du--Bentler `UG` option
and the continuous-data `RLS` statistic belong to a different setting, so test
names here do not use either suffix.

```{r setup}
library("semTests")
library("lavaan")

model <- "
  visual  =~ x1 + x2 + x3
  textual =~ x4 + x5 + x6
  speed   =~ x7 + x8 + x9
"

ordered_names <- paste0("x", 1:9)
ordinal_data <- HolzingerSwineford1939
ordinal_data[ordered_names] <- lapply(
  ordinal_data[ordered_names],
  function(x) ordered(cut(
    x,
    breaks = c(-Inf, quantile(x, c(.33, .67)), Inf),
    include.lowest = TRUE
  ))
)
```

The cuts create a tidy teaching example. They are not a recommendation to turn
continuous measurements into categories.

## Overall goodness of fit

lavaan's `WLSMV` shortcut fits the DWLS family and requests a
mean-and-variance-adjusted test. From there, `pvalues()` works as usual:

```{r ordinal-fit}
fit_ordinal <- cfa(
  model, ordinal_data,
  ordered = ordered_names,
  estimator = "WLSMV"
)

pvalues(fit_ordinal, c("STD", "SB", "SS", "ALL", "PEBA4"))
```

The footer records both the base estimator (`DWLS`) and the estimator requested
from lavaan (`WLSMV`). It also records the categorical parameterization.

```{r ordinal-provenance}
attr(pvalues(fit_ordinal, "PEBA4"), "semtests")
```

The `SB` and `SS` results give us a handy sanity check against lavaan's public
robust corrections. The eigenvalue methods use more of the reference
distribution, so they need not give the same p-value.

## Estimator families

The accepted shortcuts include:

* DWLS, WLSM, WLSMV, and WLSMVS
* ULS, ULSM, and ULSMV

Full WLS/ADF is refused. Its weight already produces an identity spectrum, so
every correction reduces to the ordinary chi-square. There is no extra
correction for `semTests` to make.

## Mixed indicators

Only variables named in `ordered` are treated as categorical. The remaining
indicators stay continuous:

```{r mixed-fit}
mixed_names <- paste0("x", 1:6)
mixed_data <- HolzingerSwineford1939
mixed_data[mixed_names] <- lapply(
  mixed_data[mixed_names],
  function(x) ordered(cut(
    x,
    breaks = c(-Inf, quantile(x, c(.33, .67)), Inf),
    include.lowest = TRUE
  ))
)

fit_mixed <- cfa(
  model, mixed_data,
  ordered = mixed_names,
  estimator = "WLSMV"
)
pvalues(fit_mixed, "PEBA4")
```

Mixed models combine thresholds with the usual covariance and mean information.
Lavaan has already done that bookkeeping, and `semTests` reads the resulting
spectrum directly.

## Pairwise missingness

With pairwise estimation, lavaan uses the observations available for each
statistic:

```{r pairwise-fit}
pairwise_data <- ordinal_data
set.seed(314)
for (variable in ordered_names) {
  pairwise_data[sample(nrow(pairwise_data), 35), variable] <- NA
}

fit_pairwise <- cfa(
  model, pairwise_data,
  ordered = ordered_names,
  estimator = "WLSMV",
  missing = "pairwise"
)
pvalues(fit_pairwise, c("SB", "SS", "PEBA4"))
```

`semTests` follows lavaan's pairwise sample-statistic and `UGamma`
calculation. Pairwise deletion brings its own assumptions about missingness, so
report and defend that choice separately. It is not FIML.

## Nested categorical models

For categorical models, nested comparison uses Satorra's 2000 construction and
the delta restriction map:

```{r categorical-nested}
constrained <- "
  visual  =~ x1 + a*x2 + a*x3
  textual =~ x4 + b*x5 + b*x6
  speed   =~ x7 + x8 + x9
"

m1 <- cfa(
  model, ordinal_data,
  ordered = ordered_names, estimator = "WLSMV"
)
m0 <- cfa(
  constrained, ordinal_data,
  ordered = ordered_names, estimator = "WLSMV"
)

pvalues_nested(
  m0, m1,
  method = "2000", A.method = "delta",
  tests = c("SB", "SS", "PALL")
)
```

Both fits must use the same requested estimator, parameterization, variables,
groups, raw data, and missingness mask. Method 2001 and the exact restriction
map are unavailable for this family.

## Practical checks

Before reading too much into a p-value, check the basics:

1. Verify convergence and adequate category counts.
2. Look for sparse categories and, when relevant, small pairwise counts.
3. Confirm that the two nested fits use identical data and options.
4. Compare `SB` and `SS` with lavaan as a quick numerical check.
5. Report the estimator and any use of pairwise missingness.

## References

Satorra, A. (2000). Scaled and adjusted restricted tests in multi-sample
analysis of moment structures. In R. D. H. Heijmans, D. S. G. Pollock, &
A. Satorra (Eds.), *Innovations in Multivariate Statistical Analysis*
(pp. 233–247). Kluwer Academic.
<https://doi.org/10.1007/978-1-4615-4603-0_17>
