---
title: "Categorical and mixed-indicator models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Categorical and mixed-indicator models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

## Scope

Categorical support is experimental. It covers lavaan's DWLS, ULS, and full-WLS
families for ordered or mixed indicators, with one or several groups and with
listwise or pairwise missingness.

The correction uses the unscaled estimator statistic and the biased spectrum
that lavaan exposes through `lavInspect(fit, "UGamma")`. The Du--Bentler `UG`
gamma and the continuous-data `RLS` statistic do not apply.

```{r setup}
library("semTests")
library("lavaan")

model <- "
  visual  =~ x1 + x2 + x3
  textual =~ x4 + x5 + x6
  speed   =~ x7 + x8 + x9
"

ordered_names <- paste0("x", 1:9)
ordinal_data <- HolzingerSwineford1939
ordinal_data[ordered_names] <- lapply(
  ordinal_data[ordered_names],
  function(x) ordered(cut(
    x,
    breaks = c(-Inf, quantile(x, c(.33, .67)), Inf),
    include.lowest = TRUE
  ))
)
```

## Overall goodness of fit

lavaan's `WLSMV` shortcut fits the DWLS family and requests a
mean-and-variance-adjusted test:

```{r ordinal-fit}
fit_ordinal <- cfa(
  model, ordinal_data,
  ordered = ordered_names,
  estimator = "WLSMV"
)

pvalues(fit_ordinal, c("STD", "SB", "SS", "ALL", "PEBA4"))
```

The output provenance records both the base estimator (`DWLS`) and the
requested estimator (`WLSMV`), together with the categorical parameterization.

```{r ordinal-provenance}
attr(pvalues(fit_ordinal, "PEBA4"), "semtests")
```

For `SB` and `SS`, the result can be sanity-checked against lavaan's public
robust corrections. The other methods use more of the spectrum than the trace
and squared trace used by these classical corrections.

## Estimator families

The accepted shortcuts include:

* DWLS, WLSM, WLSMV, and WLSMVS;
* ULS, ULSM, and ULSMV;
* full WLS.

Full WLS is a useful edge case: its weight is already the inverse asymptotic
moment covariance, so the spectrum is the identity and every correction reduces
to the ordinary chi-square test.

## Mixed indicators

Only variables named in `ordered` are treated as categorical. The remaining
indicators stay continuous:

```{r mixed-fit}
mixed_names <- paste0("x", 1:6)
mixed_data <- HolzingerSwineford1939
mixed_data[mixed_names] <- lapply(
  mixed_data[mixed_names],
  function(x) ordered(cut(
    x,
    breaks = c(-Inf, quantile(x, c(.33, .67)), Inf),
    include.lowest = TRUE
  ))
)

fit_mixed <- cfa(
  model, mixed_data,
  ordered = mixed_names,
  estimator = "WLSMV"
)
pvalues(fit_mixed, "PEBA4")
```

The threshold, covariance, and mean components of mixed sample statistics make
the moment space larger than in a continuous covariance-only analysis. Using
lavaan's inspected spectrum avoids reconstructing these conventions inside
`semTests`.

## Pairwise missingness

Pairwise categorical estimation is not FIML. It tells lavaan to estimate each
sample statistic from the observations available for that pair:

```{r pairwise-fit}
pairwise_data <- ordinal_data
set.seed(314)
for (variable in ordered_names) {
  pairwise_data[sample(nrow(pairwise_data), 35), variable] <- NA
}

fit_pairwise <- cfa(
  model, pairwise_data,
  ordered = ordered_names,
  estimator = "WLSMV",
  missing = "pairwise"
)
pvalues(fit_pairwise, c("SB", "SS", "PEBA4"))
```

This support reproduces lavaan's pairwise sample-statistic and `UGamma`
calculation. It is not a claim that pairwise deletion is generally valid under
MAR. Missingness assumptions should be justified separately.

## Nested categorical models

Categorical comparison uses Satorra's 2000 restriction-map construction with
the delta method:

```{r categorical-nested}
constrained <- "
  visual  =~ x1 + a*x2 + a*x3
  textual =~ x4 + b*x5 + b*x6
  speed   =~ x7 + x8 + x9
"

m1 <- cfa(
  model, ordinal_data,
  ordered = ordered_names, estimator = "WLSMV"
)
m0 <- cfa(
  constrained, ordinal_data,
  ordered = ordered_names, estimator = "WLSMV"
)

pvalues_nested(
  m0, m1,
  method = "2000", A.method = "delta",
  tests = c("SB", "SS", "PALL")
)
```

Both fits must use the same requested estimator, parameterization, variables,
groups, raw data, and missingness mask. Method 2001 and the exact restriction
map are intentionally rejected for this family.

## Practical checks

Before interpreting the result:

1. verify convergence and adequate category counts;
2. inspect sparse thresholds and pairwise observation counts;
3. confirm that the two nested fits use identical data and options;
4. compare `SB` and `SS` with lavaan as a trace-level sanity check;
5. treat pairwise missingness and the experimental estimator surface as
   analysis assumptions, not implementation details.

## References

Satorra, A. (2000). Scaled and adjusted restricted tests in multi-sample
analysis of moment structures. In R. D. H. Heijmans, D. S. G. Pollock, &
A. Satorra (Eds.), *Innovations in Multivariate Statistical Analysis*
(pp. 233–247). Kluwer Academic.
<https://doi.org/10.1007/978-1-4615-4603-0_17>
