---
title: "Complete continuous-data models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Complete continuous-data models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

## Scope

This vignette covers models with continuous, complete data. Normal-theory
maximum likelihood is the mature, paper-backed core of `semTests`. Complete-data
GLS and ULS use the same minimum-discrepancy framework but remain experimental
in this release.

```{r setup}
library("semTests")
library("lavaan")

model <- "
  visual  =~ x1 + x2 + x3
  textual =~ x4 + x5 + x6
  speed   =~ x7 + x8 + x9
"
data <- HolzingerSwineford1939
```

## Overall goodness of fit

Fit with a robust lavaan test so the fitted object contains the asymptotic
moment covariance needed by the eigenvalue corrections.

```{r ml-fit}
fit_ml <- cfa(model, data, estimator = "MLM")
pvalues(fit_ml)
```

The default uses penalized eigenvalue block averaging. Several corrections can
be requested together:

```{r ml-battery}
pvalues(
  fit_ml,
  c("STD_ML", "SB_ML", "SS_ML", "ALL_ML", "PEBA4_ML")
)
```

The suffixes select three separate ingredients:

* the approximation to the weighted chi-square law (`SB`, `SS`, `ALL`,
  `PEBA4`, and others);
* biased gamma or the unbiased Du--Bentler estimate (`UG`);
* the standard discrepancy statistic (`ML`) or Browne's reweighted statistic
  (`RLS`).

The unbiased gamma and RLS statistic are specific to this classical
complete-data ML setting:

```{r classical-options}
pvalues(
  fit_ml,
  c("SB_UG_ML", "PEBA4_UG_ML", "PEBA4_RLS", "PEBA4_UG_RLS")
)
```

## Reading the result

The printed footer records what was actually used. The same information is
available programmatically:

```{r provenance}
result <- pvalues(fit_ml, "PEBA4_ML")
attr(result, "semtests")
```

This is useful when results from several estimator families are collected in
one analysis.

## Nested comparison

The constrained model goes first. Here two loading equalities are tested
jointly:

```{r nested-fit}
constrained <- "
  visual  =~ x1 + a*x2 + a*x3
  textual =~ x4 + b*x5 + b*x6
  speed   =~ x7 + x8 + x9
"

m1 <- cfa(model, data, estimator = "MLM")
m0 <- cfa(constrained, data, estimator = "MLM")

pvalues_nested(m0, m1)
```

Satorra's 2000 reduction is the default. The complete-data family also supports
the 2001 construction:

```{r nested-methods}
pvalues_nested(m0, m1, method = "2000", tests = c("SB_ML", "PALL_ML"))
pvalues_nested(m0, m1, method = "2001", tests = c("SB_ML", "PALL_ML"))
```

`PALL` is the recommended default for nested comparison. The two reductions
need not agree in finite samples because they construct the restriction
spectrum differently.

## GLS and ULS

The limiting weighted chi-square argument is not unique to ML. GLS supplies
the required covariance directly; ULS should be fit with a robust test so
lavaan computes it:

```{r least-squares}
fit_gls <- cfa(model, data, estimator = "GLS")
fit_uls <- cfa(
  model, data,
  estimator = "ULS", test = "satorra.bentler"
)

pvalues(fit_gls, "PEBA4")
pvalues(fit_uls, "PEBA4")
```

For GLS and ULS, use the standard statistic and biased gamma. Requests for
`RLS` or `UG` are rejected because their current derivations belong to the
classical ML case.

## Practical checks

For a complete-data analysis:

1. verify that both lavaan fits converged and use the same data;
2. fit a robust test whenever lavaan does not otherwise expose `Gamma`;
3. report the test name, base statistic, and gamma choice;
4. retain the `semtests` provenance attribute with saved results;
5. use the nested methods only for genuinely nested models.

See `vignette("categorical-data", package = "semTests")` and
`vignette("fiml-missing-data", package = "semTests")` before carrying this
workflow to categorical or missing data.
