---
title: "FIML models and information conventions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{FIML models and information conventions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

## Scope

`semTests` supports continuous, single-group FIML models without fixed
exogenous covariates. FIML uses the standard likelihood-ratio statistic and the
biased gamma estimate. The complete-data `RLS` and Du--Bentler `UG` choices are
not defined for this family.

```{r setup}
library("semTests")
library("lavaan")

model <- "
  visual  =~ x1 + x2 + x3
  textual =~ x4 + x5 + x6
  speed   =~ x7 + x8 + x9
"
```

## A reproducible MAR-like example

To make the information choice visible, missingness in `x1` and `x5` is made
dependent on `x9`, which remains observed. This is an illustrative mechanism,
not a missingness diagnosis for the original data.

```{r missing-data}
fiml_data <- HolzingerSwineford1939
set.seed(20260717)
driver <- as.numeric(scale(fiml_data$x9))

missing_x1 <- runif(nrow(fiml_data)) <
  plogis(qlogis(.20) + .8 * driver)
missing_x5 <- runif(nrow(fiml_data)) <
  plogis(qlogis(.20) - .8 * driver)

fiml_data$x1[missing_x1] <- NA
fiml_data$x5[missing_x5] <- NA
colMeans(is.na(fiml_data[paste0("x", 1:9)]))
```

Fit the model using FIML:

```{r fiml-fit}
fit_observed <- cfa(
  model, fiml_data,
  missing = "fiml",
  estimator = "MLR"
)

pvalues(fit_observed, c("SB", "SS", "PEBA4"))
```

## Two information choices that should not be conflated

There are two related but distinct controls:

1. lavaan's `information` fit option selects the model information stored and
   used by lavaan's own inference machinery;
2. `fiml.convention` selects the matrices used to construct the semTests
   reference spectrum.

The semTests default, `"observed"`, uses observed saturated and observed model
information throughout. It requests these matrices explicitly, so its result
does not silently change because the lavaan fit was created with expected
information.

The `"lavaan"` convention is a compatibility construction. For a single model
it reproduces `lavInspect(fit, "UGamma")`; for a nested model it reproduces
lavaan's public Satorra-2000 construction. That construction uses lavaan's H1
weight convention and the model information selected in the fit. It is
therefore not simply another spelling of `information = "expected"`.

## Why observed information is the default

The theoretical case is established: expected information can be inconsistent
under general MAR, while observed information reflects the realized
missingness patterns and sample curvature (Kenward & Molenberghs, 1998;
Savalei, 2010; Savalei & Rosseel, 2022). A focused computational study was used
to check whether that choice also matters in the finite samples relevant here.

The study crossed:

* multivariate normal and elliptical \(t_5\) data;
* a correct two-factor CFA and a model omitting a 0.25 cross-loading;
* complete data, 30% MCAR, and 30% MAR;
* \(N = 150, 300, 600,\) and \(1200\).

Each of the 48 cells used 200,000 replications, for 9.6 million fits. A free
loading and the factor covariance were followed. The same fitted model was
used to compare expected information, the analytic observed-H1 approximation,
and the full observed Hessian. Each information matrix was paired with either
its model-based inverse or the same empirical-score sandwich meat (Yuan &
Bentler, 2000).

```{r information-study, echo = FALSE}
information_coverage <- data.frame(
  Method = c(
    "Expected, model-based",
    "Expected + sandwich",
    "Observed H1, model-based",
    "Observed H1 + sandwich",
    "Full observed Hessian, model-based",
    "Full observed Hessian + sandwich"
  ),
  `Correct, normal` = c(94.48, 94.13, 94.67, 94.39, 94.73, 94.72),
  `Correct, t5` = c(79.41, 92.45, 79.98, 92.52, 80.14, 93.64),
  `Misspecified, normal` = c(92.10, 91.53, 92.72, 92.46, 93.46, 94.53),
  `Misspecified, t5` = c(75.21, 89.28, 76.38, 89.98, 77.51, 93.00),
  check.names = FALSE
)
knitr::kable(
  information_coverage,
  digits = 2,
  caption = paste(
    "Mean 95% coverage across sample sizes, missingness mechanisms,",
    "and focal parameters. Misspecified cells use centered coverage."
  )
)
```

Across correctly specified cells, the full observed Hessian with sandwich
covariance had the smallest mean absolute coverage error, 0.821 percentage
points, and mean coverage of 94.179%. Under a correct normal model the two
full-Hessian covariances were essentially tied near 94.7%. Under \(t_5\) data,
model-based covariance covered only about 80%, while the full-Hessian sandwich
reached 93.64%. Under misspecification, its centered coverage was 94.53% for
normal data and 93.00% for \(t_5\), compared with 92.46% and 89.98% for the
observed-H1 sandwich.

Rare converged Heywood or nearly singular solutions produced extreme
estimate and standard-error tails, so arithmetic mean SE/SD ratios were not
used. Coverage retained converged replications with finite standard errors,
and the worst cell-specific information-inversion rate was 0.0105%. Across all
48 design cells, the simulation implementation and lavaan agreed in 576 of 576
parameter-level standard-error comparisons, with maximum absolute difference
\(5.81\times10^{-7}\).

This study concerns parameter standard errors, not the null calibration of
`semTests` goodness-of-fit *p*-values. It supports using observed rather than
expected information as the default building block, especially when the full
observed model information enters a nested comparison. Deterministic spectrum
parity and the seeded null simulation under `tools/` provide the separate
checks of the `semTests` testing pipeline. The `"lavaan"` convention remains
available when reproduction of lavaan's inspected spectrum is the goal.

## Comparing observed and expected model information

Fit the same model again, changing only lavaan's model-information option:

```{r expected-fit}
fit_expected <- cfa(
  model, fiml_data,
  missing = "fiml",
  estimator = "MLR",
  information = "expected"
)

max(abs(coef(fit_observed) - coef(fit_expected)))
```

The option does not change the fitted parameters here. It can change the
reference spectrum:

```{r convention-table}
tests <- c("SB", "SS", "PEBA4")

comparison <- cbind(
  observed_fit__observed_convention = as.numeric(
    pvalues(fit_observed, tests, fiml.convention = "observed")
  ),
  observed_fit__lavaan_convention = as.numeric(
    pvalues(fit_observed, tests, fiml.convention = "lavaan")
  ),
  expected_fit__observed_convention = as.numeric(
    pvalues(fit_expected, tests, fiml.convention = "observed")
  ),
  expected_fit__lavaan_convention = as.numeric(
    pvalues(fit_expected, tests, fiml.convention = "lavaan")
  )
)
rownames(comparison) <- tests
format(comparison, scientific = TRUE, digits = 4)
```

The two observed-convention columns agree: semTests explicitly uses observed
information. The lavaan-convention columns can differ because the compatibility
path honors the model-information choice stored in the fit.

Observed information reflects the realized missingness patterns and sample
curvature. Expected information averages curvature under the fitted model.
They can be close under favorable conditions, but missingness and non-normality
make it unwise to assume they are interchangeable. `semTests` therefore makes
the convention visible in both the function call and the result provenance.

## Nested FIML comparison

For nested models, both fits must use the same raw observations and missingness
mask:

```{r nested-fits}
constrained <- "
  visual  =~ x1 + a*x2 + a*x3
  textual =~ x4 + b*x5 + b*x6
  speed   =~ x7 + x8 + x9
"

m1 <- cfa(
  model, fiml_data,
  missing = "fiml", estimator = "MLR"
)
m0 <- cfa(
  constrained, fiml_data,
  missing = "fiml", estimator = "MLR"
)
```

The default delta map compares the local moment-tangent spaces. The exact map
uses the literal parameter restriction and requires the same underlying full
free-parameter set:

```{r nested-comparison}
nested_tests <- c("SB", "SS", "PEBA2")
nested <- cbind(
  observed_delta = as.numeric(pvalues_nested(
    m0, m1, tests = nested_tests,
    A.method = "delta", fiml.convention = "observed"
  )),
  observed_exact = as.numeric(pvalues_nested(
    m0, m1, tests = nested_tests,
    A.method = "exact", fiml.convention = "observed"
  )),
  lavaan_delta = as.numeric(pvalues_nested(
    m0, m1, tests = nested_tests,
    A.method = "delta", fiml.convention = "lavaan"
  ))
)
rownames(nested) <- nested_tests
format(nested, scientific = TRUE, digits = 4)
```

Use the delta default when equivalent models have different parameter-table
representations. Use exact when the scientific hypothesis is naturally a
literal restriction on a shared free-parameter set.

## What to report

A reproducible FIML analysis should report:

* the missingness pattern and its substantive justification;
* lavaan's estimator and `information` option;
* the semTests `fiml.convention`;
* the nested `A.method`, when applicable;
* the requested p-value method and unscaled base-statistic choice.

The provenance footer and `attr(result, "semtests")` retain these semTests
choices with the returned p-values.

## References

Kenward, M. G., & Molenberghs, G. (1998). Likelihood based frequentist
inference when data are missing at random. *Statistical Science*, 13(3),
236–247. <https://doi.org/10.1214/ss/1028905886>

Savalei, V. (2010). Expected versus observed information in SEM with incomplete
normal and nonnormal data. *Psychological Methods*, 15(4), 352–367.
<https://doi.org/10.1037/a0020143>

Savalei, V., & Rosseel, Y. (2022). Computational options for standard errors
and test statistics with incomplete normal and nonnormal data in SEM.
*Structural Equation Modeling*, 29(2), 163–181.
<https://doi.org/10.1080/10705511.2021.1877548>

Satorra, A. (2000). Scaled and adjusted restricted tests in multi-sample
analysis of moment structures. In R. D. H. Heijmans, D. S. G. Pollock, &
A. Satorra (Eds.), *Innovations in Multivariate Statistical Analysis*
(pp. 233–247). Kluwer Academic.
<https://doi.org/10.1007/978-1-4615-4603-0_17>

Yuan, K.-H., & Bentler, P. M. (2000). Three likelihood-based methods for mean
and covariance structure analysis with nonnormal missing data.
*Sociological Methodology*, 30, 165–200.
<https://doi.org/10.1111/0081-1750.00078>
