---
title: "FIML models and information conventions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{FIML models and information conventions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

## Scope

`semTests` supports continuous, single-group FIML models without fixed
exogenous covariates. FIML uses the standard likelihood-ratio statistic and the
biased gamma estimate. The complete-data `RLS` and Du--Bentler `UG` choices are
not defined for this family.

```{r setup}
library("semTests")
library("lavaan")

model <- "
  visual  =~ x1 + x2 + x3
  textual =~ x4 + x5 + x6
  speed   =~ x7 + x8 + x9
"
```

## A reproducible MAR-like example

To make the information choice visible, missingness in `x1` and `x5` is made
dependent on `x9`, which remains observed. This is an illustrative mechanism,
not a missingness diagnosis for the original data.

```{r missing-data}
fiml_data <- HolzingerSwineford1939
set.seed(20260717)
driver <- as.numeric(scale(fiml_data$x9))

missing_x1 <- runif(nrow(fiml_data)) <
  plogis(qlogis(.20) + .8 * driver)
missing_x5 <- runif(nrow(fiml_data)) <
  plogis(qlogis(.20) - .8 * driver)

fiml_data$x1[missing_x1] <- NA
fiml_data$x5[missing_x5] <- NA
colMeans(is.na(fiml_data[paste0("x", 1:9)]))
```

Fit the model using FIML:

```{r fiml-fit}
fit_observed <- cfa(
  model, fiml_data,
  missing = "fiml",
  estimator = "MLR"
)

pvalues(fit_observed, c("SB", "SS", "PEBA4"))
```

## Two information choices that should not be conflated

There are two related but distinct controls:

1. lavaan's `information` fit option selects the model information stored and
   used by lavaan's own inference machinery;
2. `fiml.convention` selects the matrices used to construct the semTests
   reference spectrum.

The semTests default, `"observed"`, uses observed saturated and observed model
information throughout. It requests these matrices explicitly, so its result
does not silently change because the lavaan fit was created with expected
information.

The `"lavaan"` convention is a compatibility construction. For a single model
it reproduces `lavInspect(fit, "UGamma")`; for a nested model it reproduces
lavaan's public Satorra-2000 construction. That construction uses lavaan's H1
weight convention and the model information selected in the fit. It is
therefore not simply another spelling of `information = "expected"`.

## Comparing observed and expected model information

Fit the same model again, changing only lavaan's model-information option:

```{r expected-fit}
fit_expected <- cfa(
  model, fiml_data,
  missing = "fiml",
  estimator = "MLR",
  information = "expected"
)

max(abs(coef(fit_observed) - coef(fit_expected)))
```

The option does not change the fitted parameters here. It can change the
reference spectrum:

```{r convention-table}
tests <- c("SB", "SS", "PEBA4")

comparison <- cbind(
  observed_fit__observed_convention = as.numeric(
    pvalues(fit_observed, tests, fiml.convention = "observed")
  ),
  observed_fit__lavaan_convention = as.numeric(
    pvalues(fit_observed, tests, fiml.convention = "lavaan")
  ),
  expected_fit__observed_convention = as.numeric(
    pvalues(fit_expected, tests, fiml.convention = "observed")
  ),
  expected_fit__lavaan_convention = as.numeric(
    pvalues(fit_expected, tests, fiml.convention = "lavaan")
  )
)
rownames(comparison) <- tests
format(comparison, scientific = TRUE, digits = 4)
```

The two observed-convention columns agree: semTests explicitly uses observed
information. The lavaan-convention columns can differ because the compatibility
path honors the model-information choice stored in the fit.

Observed information reflects the realized missingness patterns and sample
curvature. Expected information averages curvature under the fitted model.
They can be close under favorable conditions, but missingness and non-normality
make it unwise to assume they are interchangeable. `semTests` therefore makes
the convention visible in both the function call and the result provenance.

## Nested FIML comparison

For nested models, both fits must use the same raw observations and missingness
mask:

```{r nested-fits}
constrained <- "
  visual  =~ x1 + a*x2 + a*x3
  textual =~ x4 + b*x5 + b*x6
  speed   =~ x7 + x8 + x9
"

m1 <- cfa(
  model, fiml_data,
  missing = "fiml", estimator = "MLR"
)
m0 <- cfa(
  constrained, fiml_data,
  missing = "fiml", estimator = "MLR"
)
```

The default delta map compares the local moment-tangent spaces. The exact map
uses the literal parameter restriction and requires the same underlying full
free-parameter set:

```{r nested-comparison}
nested <- cbind(
  observed_delta = as.numeric(pvalues_nested(
    m0, m1, tests = tests,
    A.method = "delta", fiml.convention = "observed"
  )),
  observed_exact = as.numeric(pvalues_nested(
    m0, m1, tests = tests,
    A.method = "exact", fiml.convention = "observed"
  )),
  lavaan_delta = as.numeric(pvalues_nested(
    m0, m1, tests = tests,
    A.method = "delta", fiml.convention = "lavaan"
  ))
)
rownames(nested) <- tests
format(nested, scientific = TRUE, digits = 4)
```

Use the delta default when equivalent models have different parameter-table
representations. Use exact when the scientific hypothesis is naturally a
literal restriction on a shared free-parameter set.

## What to report

A reproducible FIML analysis should report:

* the missingness pattern and its substantive justification;
* lavaan's estimator and `information` option;
* the semTests `fiml.convention`;
* the nested `A.method`, when applicable;
* the requested p-value method and the unscaled base statistic.

The provenance footer and `attr(result, "semtests")` retain these semTests
choices with the returned p-values.
