---
title: "Measurement invariance with ordinal indicators"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Measurement invariance with ordinal indicators}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

## Scope

Measurement invariance asks whether a construct is measured the same way across
groups. The usual strategy fits a sequence of increasingly constrained models
and compares each model with the one before it.

Ordinal indicators need a little care here. Thresholds enter the measurement
model alongside factor loadings, and the order of the invariance tests matters.
Following Wu and Estabrook (2016), we test threshold equality first and loading
equality second. The example is deliberately small, but it gives
`pvalues_nested()` a realistic categorical job to do.

```{r setup}
library("semTests")
library("lavaan")

model <- "
  visual  =~ x1 + x2 + x3
  textual =~ x4 + x5 + x6
  speed   =~ x7 + x8 + x9
"

ordered_names <- paste0("x", 1:9)
mi_data <- HolzingerSwineford1939
mi_data[ordered_names] <- lapply(
  mi_data[ordered_names],
  function(x) ordered(cut(
    x,
    breaks = c(-Inf, quantile(x, c(.25, .50, .75)), Inf),
    include.lowest = TRUE
  ))
)
```

The grouping variable is `school`, which has two levels. Cutting at the
quartiles gives every indicator four ordered categories and three thresholds.
This artificial split is only here to make the example reproducible. Ordinal
status should come from the measurement design in a real study.

## The configural model

The configural model uses the same factor pattern in both schools while leaving
its measurement parameters free across schools. We can start by checking its
overall fit:

```{r configural}
configural <- cfa(
  model, mi_data, ordered = ordered_names,
  group = "school", estimator = "WLSMV"
)

pvalues(configural, c("SB", "SS", "ALL", "PEBA4"))
```

## Thresholds first

The next model makes the thresholds equal across schools. It is more
constrained than the configural model, so it goes in the first argument:

```{r thresholds}
equal_thresholds <- cfa(
  model, mi_data, ordered = ordered_names,
  group = "school", group.equal = "thresholds", estimator = "WLSMV"
)

pvalues_nested(
  equal_thresholds, configural,
  tests = c("SB", "SS", "ALL", "PEBA4")
)
```

The test asks whether the equal-threshold restrictions create a noticeable loss
of fit. A large p-value means the data have not picked a fight with those
restrictions. It does not prove that the thresholds are identical.

## Then the loadings

Now retain equal thresholds and add equality constraints on the loadings:

```{r loadings}
equal_loadings <- cfa(
  model, mi_data, ordered = ordered_names,
  group = "school",
  group.equal = c("thresholds", "loadings"),
  estimator = "WLSMV"
)

pvalues_nested(
  equal_loadings, equal_thresholds,
  tests = c("SB", "SS", "ALL", "PEBA4")
)
```

This second comparison isolates the extra loading restrictions given equal
thresholds. A small p-value would count against that step. A large one says the
step did not noticeably worsen fit in this sample. In a real invariance study,
the model-identification choices, effect sizes, and the purpose of the group
comparison deserve attention too.

Both comparisons use Satorra's 2000 reduction and the delta restriction map.
Those are the categorical nested options supported by `semTests`.

```{r provenance}
attr(
  pvalues_nested(
    equal_loadings, equal_thresholds,
    tests = "PEBA4"
  ),
  "semtests"
)
```

The result footer records the categorical estimator, the reduction method, and
the restriction map. That makes it much easier to tell later which test you
actually ran.

## References

Satorra, A. (2000). Scaled and adjusted restricted tests in multi-sample
analysis of moment structures. In R. D. H. Heijmans, D. S. G. Pollock, &
A. Satorra (Eds.), *Innovations in Multivariate Statistical Analysis*
(pp. 233–247). Kluwer Academic.
<https://doi.org/10.1007/978-1-4615-4603-0_17>

Wu, H., & Estabrook, R. (2016). Identification of confirmatory factor analysis
models of different levels of invariance for ordered categorical outcomes.
*Psychometrika*, 81(4), 1014--1045.
<https://doi.org/10.1007/s11336-016-9506-0>
