---
title: "Measurement invariance with ordinal indicators"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Measurement invariance with ordinal indicators}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

## Scope

Measurement invariance asks whether a construct is measured the same way across
groups. It is tested as a chain of nested models: a *configural* model that only
shares the factor pattern, then a *metric* model that additionally holds the
measurement parameters equal across groups. Each step is a nested-model
comparison, and with ordinal indicators the difference statistic is a weighted
sum of chi-squares -- exactly the setting the `semTests` robust p-values are
built for. This vignette runs the sequence on ordinal data and reports the
block-averaged and penalized tests next to the Satorra--Bentler value.

```{r setup}
library("semTests")
library("lavaan")

model <- "
  visual  =~ x1 + x2 + x3
  textual =~ x4 + x5 + x6
  speed   =~ x7 + x8 + x9
"

ordered_names <- paste0("x", 1:9)
mi_data <- HolzingerSwineford1939
mi_data[ordered_names] <- lapply(
  mi_data[ordered_names],
  function(x) ordered(cut(
    x,
    breaks = c(-Inf, quantile(x, c(.33, .67)), Inf),
    include.lowest = TRUE
  ))
)
```

The grouping variable is `school`, which has two levels.

## The configural model

The configural model estimates the measurement parameters freely in each group.
Its overall fit is the usual single-model goodness-of-fit question:

```{r configural}
configural <- cfa(
  model, mi_data, ordered = ordered_names,
  group = "school", estimator = "WLSMV"
)

pvalues(configural, c("SB", "SS", "ALL", "PEBA4"))
```

## The metric invariance test

Metric invariance holds the measurement model equal across groups. For ordinal
indicators the loadings and thresholds are identified together (Wu and
Estabrook, 2016), so `group.equal = "loadings"` constrains the measurement model
as a whole. The constrained model is the *more* restricted one, so it goes in
`m0`:

```{r metric}
metric <- cfa(
  model, mi_data, ordered = ordered_names,
  group = "school", group.equal = "loadings", estimator = "WLSMV"
)

pvalues_nested(metric, configural, tests = c("SB", "SS", "ALL", "PEBA4"))
```

A large p-value supports invariance; a small one says the measurement model
differs across groups. The comparison uses Satorra's 2000 reduction with the
delta restriction map, which `semTests` selects automatically for ordinal fits
(the only supported categorical nested configuration).

```{r provenance}
attr(pvalues_nested(metric, configural, tests = "PEBA4"), "semtests")
```

The provenance footer records the categorical estimator, the reduction method,
and the restriction map, so the inferential choices stay attached to the result.

## Reference

Wu, H., & Estabrook, R. (2016). Identification of confirmatory factor analysis
models of different levels of invariance for ordered categorical outcomes.
*Psychometrika*, 81(4), 1014--1045.
