---
title: "Robust test statistics with semTests"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Robust test statistics with semTests}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

## Introduction

`semTests` computes robust *p*-values for structural equation models fit with
[`lavaan`](https://lavaan.ugent.be/). The standard chi-square test of model fit
can reject too often when the data are non-normal. Familiar corrections such as
Satorra--Bentler help. The methods in this package keep more information about
the shape of the reference distribution by using its estimated eigenvalues.

For the mathematically curious, the reference distribution is a weighted sum
\(\sum_j \lambda_j \chi^2_1\). You do not need to calculate any eigenvalues by
hand. Fit the model in lavaan, then let `pvalues()` do that part. The same basic
idea covers maximum likelihood, least squares, categorical models, and FIML.

The penalized eigenvalue block-averaging (`pEBA`) and penalized regression
(`pOLS`) methods were introduced by Foldnes, Moss, and Grønneberg (2025).
Foldnes, Grønneberg, and Moss (2026) extended them to nested model comparison.

```{r setup}
library("semTests")
library("lavaan")
```

## Choose a focused guide

This vignette is a quick tour. Pick the focused guide that matches your
data when you want more depth:

* `vignette("continuous-data", package = "semTests")` covers complete
  continuous data, the classical ML workflow, GLS/ULS, and nested tests.
* `vignette("categorical-data", package = "semTests")` covers ordered and mixed
  indicators, pairwise missingness, and categorical nested comparison.
* `vignette("fiml-missing-data", package = "semTests")` covers continuous FIML,
  the finite-sample evidence for the observed-information default, the lavaan
  compatibility convention, and nested restriction maps.
* `vignette("latent-growth", package = "semTests")` gives FIML a more applied
  outing in a latent growth model.
* `vignette("measurement-invariance", package = "semTests")` walks through
  threshold and loading invariance with ordinal indicators.

## Basic usage: goodness of fit

Fit a model with `lavaan`, then call `pvalues()`. For ML models, request a
robust lavaan test such as `estimator = "MLM"` or `"MLR"`. This makes lavaan
store the extra sample information that `semTests` needs.

```{r basic}
model <- "visual  =~ x1 + x2 + x3
          textual =~ x4 + x5 + x6
          speed   =~ x7 + x8 + x9"
fit <- cfa(model, HolzingerSwineford1939, estimator = "MLM")
pvalues(fit)
```

The result is a named vector with a short footer showing the estimator, data
type, information choice, and degrees of freedom actually used. The complete
record is available with `attr(x, "semtests")`.

### The test-name grammar

Each requested test uses one of `TEST`, `TEST_UG`, `TEST_ML`, `TEST_RLS`,
`TEST_UG_ML`, or `TEST_UG_RLS`:

* **TEST** gives the family. Choose `pEBA<j>` for penalized block averaging,
  where `j` is usually between 2 and 6. Other choices include
  `pOLS<gamma>`, `pall`, `all`, `eba<j>`, `std`, `sb`, `ss`, and `sf`.
* **UG** asks for the unbiased Du--Bentler (2022) gamma. Without `UG`, the
  standard biased gamma is used.
* **ML / RLS** chooses the base chi-square. `ML` uses the normal-theory
  discrepancy and `RLS` uses Browne's reweighted least squares statistic.
  Leave the suffix out to use the default for the fitted model.

Several tests can be requested at once:

```{r grammar}
pvalues(fit, c("SB_ML", "pEBA4_ML", "pOLS2_ML"))
```

The `UG` gamma and the `RLS` statistic belong to continuous, complete-data fits
from lavaan's ML family. That includes fits requested as `MLM` or `MLR`. Other
model families return an error if either option is requested.

## Nested model comparison

To compare two nested models, fit both and pass them to `pvalues_nested()` with
the **constrained** model first. Here we ask whether the textual loadings can be
held equal:

```{r nested}
constrained <- "visual  =~ x1 + x2 + x3
                textual =~ a*x4 + a*x5 + a*x6
                speed   =~ x7 + x8 + x9"
m1 <- cfa(model, HolzingerSwineford1939, estimator = "MLM")
m0 <- cfa(constrained, HolzingerSwineford1939, estimator = "MLM")
pvalues_nested(m0, m1)
```

`PALL` is the recommended default for nested comparison (Foldnes, Grønneberg,
and Moss, 2026). `semTests` uses Satorra's 2000 construction. The 2001 method has been
withdrawn because it performs poorly. Both fits must use the same estimator, options,
variables, groups, and observations. If the inputs arrive in reverse order,
`semTests` warns and swaps them.

## More data families

The package also covers several non-ML estimators. These paths have been checked
against a separate implementation. `?semTests-support` gives the exact list of
supported combinations.

GLS and ULS work directly. ULS needs a robust lavaan test so the fit contains
the required sample information:

```{r estimators}
gls <- cfa(model, HolzingerSwineford1939, estimator = "GLS")
pvalues(gls, "pEBA4")

uls <- cfa(model, HolzingerSwineford1939, estimator = "ULS",
           test = "satorra.bentler")
pvalues(uls, "pEBA4")
```

Missing data is handled through full-information maximum likelihood. Fit with
`missing = "fiml"` and pass the fit to `pvalues()` as usual. FIML uses the
usual biased gamma and the standard statistic, so suffix-free test names are
the right choice.

```{r fiml}
HS <- HolzingerSwineford1939
set.seed(1)
HS$x1[sample(nrow(HS), 60)] <- NA
fit_fiml <- cfa(model, HS, missing = "fiml", estimator = "MLR")
pvalues(fit_fiml)
```

Observed information is the default, following Savalei (2010). The choice and
the optional lavaan-compatibility convention are explained in
`vignette("fiml-missing-data", package = "semTests")`, along with nested FIML
examples.

### Ordered and mixed indicators

Categorical models use lavaan's inspected `UGamma` and unscaled statistic.
DWLS, including WLSMV/WLSM/WLSMVS, and ULS, including ULSMV, are supported for
single-group and multigroup ordered or mixed-indicator models. Full WLS/ADF is
refused because its correction reduces to the ordinary chi-square.

```{r categorical}
HSord <- HolzingerSwineford1939
ordered_names <- paste0("x", 1:9)
HSord[ordered_names] <- lapply(
  HSord[ordered_names],
  function(x) ordered(cut(x, 3))
)
fit_ordinal <- cfa(model, HSord, ordered = ordered_names)
pvalues(fit_ordinal, c("SB", "SS", "pEBA4"))
```

The artificial cut only keeps this tour reproducible. In a real analysis, the
decision to treat an indicator as ordered should come from how it was measured.

Categorical fits may use listwise or pairwise missingness. Pairwise support
uses lavaan's pairwise sample statistics and `UGamma`. Pairwise deletion has
its own missingness assumptions and does not provide FIML inference. Nested
categorical comparison uses `method = "2000"` with `A.method = "delta"`.

## References

Browne, M. W. (1974). Generalized least squares estimators in the analysis of
covariance structures. *South African Statistical Journal*, 8, 1–24.

Du, H., & Bentler, P. M. (2022). 40-Year Old Unbiased Distribution Free
Estimator Reliably Improves SEM Statistics for Nonnormal Data. *Structural
Equation Modeling*, 29(6), 872–887.
<https://doi.org/10.1080/10705511.2022.2063870>

Foldnes, N., Moss, J., & Grønneberg, S. (2025). Improved goodness of fit
procedures for structural equation models. *Structural Equation Modeling: A
Multidisciplinary Journal*, 32(1), 1–13.
<https://doi.org/10.1080/10705511.2024.2372028>

Foldnes, N., Grønneberg, S., & Moss, J. (2026). Penalized eigenvalue block
averaging: Extension to nested model comparison and Monte Carlo evaluations.
*Behavior Research Methods*, 58, article 107.
<https://doi.org/10.3758/s13428-026-02968-4>

Savalei, V. (2010). Expected versus observed information in SEM with incomplete
normal and nonnormal data. *Psychological Methods*, 15(4), 352–367.
<https://doi.org/10.1037/a0020143>

Satorra, A. (2000). Scaled and adjusted restricted tests in multi-sample
analysis of moment structures. In R. D. H. Heijmans, D. S. G. Pollock, &
A. Satorra (Eds.), *Innovations in Multivariate Statistical Analysis*
(pp. 233–247). Kluwer Academic.
<https://doi.org/10.1007/978-1-4615-4603-0_17>

Satorra, A., & Bentler, P. M. (1994). Corrections to test statistics and
standard errors in covariance structure analysis. In A. von Eye & C. C. Clogg
(Eds.), *Latent Variables Analysis: Applications for Developmental Research*
(pp. 399–419). Sage.
