---
title: "Robust test statistics with semTests"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Robust test statistics with semTests}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
options(warn = -1)
```

## Introduction

`semTests` computes robust *p*-values for structural equation models fit with
[`lavaan`](https://lavaan.ugent.be/). The standard chi-square test of model fit
is anti-conservative under non-normality, and the usual corrections
(Satorra–Bentler and friends) only partly fix it. The eigenvalue-based tests in
this package target the *limiting null law* of the test statistic directly: that
law is a weighted sum of independent chi-squares,
\(\sum_j \lambda_j \chi^2_1\), whose weights \(\lambda_j\) are the eigenvalues of
a matrix built from the model and the asymptotic covariance of the sample
moments. Because this holds for any minimum-discrepancy estimator, the tests are
not specific to normal-theory maximum likelihood.

The penalized eigenvalue block-averaging (`pEBA`) and penalized regression
(`pOLS`) methods were introduced by Foldnes, Moss, & Grønneberg (2025) and
extended to nested model comparison by Foldnes, Grønneberg, & Moss (2026).

```{r setup}
library("semTests")
library("lavaan")
```

## Choose a focused guide

This vignette is the short overview. Three focused vignettes develop the
supported families separately:

* `vignette("continuous-data", package = "semTests")` covers complete
  continuous data, the classical ML workflow, GLS/ULS, and nested tests.
* `vignette("categorical-data", package = "semTests")` covers ordered and mixed
  indicators, pairwise missingness, and categorical nested comparison.
* `vignette("fiml-missing-data", package = "semTests")` covers continuous FIML,
  the observed-information and lavaan compatibility conventions, and nested
  restriction maps.

## Basic usage: goodness of fit

Fit a model with `lavaan`, then call `pvalues()`. Fit with a robust test (e.g.
`estimator = "MLM"` or `"MLR"`) so that lavaan exposes the asymptotic moment
covariance the correction needs.

```{r basic}
model <- "visual  =~ x1 + x2 + x3
          textual =~ x4 + x5 + x6
          speed   =~ x7 + x8 + x9"
fit <- cfa(model, HolzingerSwineford1939, estimator = "MLM")
pvalues(fit)
```

The result is a named vector that also prints a one-line provenance footer
recording the estimator, data type, information matrix, and degrees of freedom
actually used. The full option record is available with
`attr(x, "semtests")`.

### The test-name grammar

Each requested test uses one of `TEST`, `TEST_UG`, `TEST_ML`, `TEST_RLS`,
`TEST_UG_ML`, or `TEST_UG_RLS`:

* **test** — the family: `pEBA<j>` (penalized block-averaging, the recommended
  default; `j` is the number of blocks, typically 2–6), `pOLS<gamma>` (penalized
  regression), `pall`/`all`, `eba<j>`, or a traditional statistic `std`, `sb`
  (Satorra–Bentler), `ss` (scaled-and-shifted), `sf` (scaled *F*).
* **ug** — include `UG` to use the unbiased Du–Bentler (2022) gamma instead of
  the standard biased one.
* **ml / rls** — the base chi-square: `ML` (normal-theory discrepancy) or `RLS`
  (Browne's reweighted least squares). When omitted, a fit-appropriate default
  is chosen.

Several tests can be requested at once:

```{r grammar}
pvalues(fit, c("SB_ML", "pEBA4_ML", "pOLS2_ML"))
```

The `UG` gamma and the `RLS` statistic are defined only for the classical case
(continuous, complete data, `estimator = "ML"`); off that case they are refused
with an informative error rather than silently degrading.

## Nested model comparison

To compare two nested models, fit both and pass them to `pvalues_nested()` with
the **constrained** model first. Here we test weak invariance of the textual
loadings:

```{r nested}
constrained <- "visual  =~ x1 + x2 + x3
                textual =~ a*x4 + a*x5 + a*x6
                speed   =~ x7 + x8 + x9"
m1 <- cfa(model, HolzingerSwineford1939, estimator = "MLM")
m0 <- cfa(constrained, HolzingerSwineford1939, estimator = "MLM")
pvalues_nested(m0, m1)
```

`pall` is the recommended default for nested comparison (Foldnes, Grønneberg, &
Moss, 2026). The reduction `method` defaults to `"2000"` (Satorra 2000);
`"2001"` is also available for complete-data fits. Both fits must use the same
estimator, fitting conventions, variables, groups, sample sizes, raw data, and
missingness mask. Reversed inputs warn before being swapped. Method 2001 stops,
rather than silently changing methods, if its leading spectrum is materially
negative.

## Other estimators and missing data (experimental)

The same machinery applies to other estimators. The non-ML paths below are
**experimental** as of this release; the classical ML path is stable. See
`?semTests-support` for the authoritative matrix of what is supported.

GLS and ULS work directly (ULS needs a robust test for the asymptotic
covariance):

```{r estimators}
gls <- cfa(model, HolzingerSwineford1939, estimator = "GLS")
pvalues(gls, "pEBA4")

uls <- cfa(model, HolzingerSwineford1939, estimator = "ULS",
           test = "satorra.bentler")
pvalues(uls, "pEBA4")
```

Missing data is handled through full-information maximum likelihood. Fit with
`missing = "fiml"` and pass the fit to `pvalues()` as usual. FIML uses the
biased gamma and the standard statistic, so the `UG`/`RLS` suffixes do not apply;
the default suffix-free tests are the right choice.

```{r fiml}
HS <- HolzingerSwineford1939
set.seed(1)
HS$x1[sample(nrow(HS), 60)] <- NA
fit_fiml <- cfa(model, HS, missing = "fiml", estimator = "MLR")
pvalues(fit_fiml)
```

The default `fiml.convention = "observed"` uses observed saturated and model
information throughout, following Kenward and Molenberghs' (1998)
observed-information recommendation, and is independently validated against
magmaan. Use
`fiml.convention = "lavaan"` when exact reproduction of lavaan 0.7-2's inspected
single-model `UGamma` or nested Satorra-2000 correction is wanted:

```{r fiml-conventions}
pvalues(fit_fiml, fiml.convention = "observed")
pvalues(fit_fiml, fiml.convention = "lavaan")
```

For nested FIML comparisons, `A.method = "delta"` is the default and works
across a wider range of equivalent parameterizations. `"exact"` uses the
literal parameter restriction map and requires the same underlying full free
parameter set in both models.

For a single FIML model, the lavaan convention is specifically
`lavInspect(fit, "UGamma")`; it need not equal the scalar
Yuan--Bentler--Mplus correction stored by a default MLR fit. For nested FIML
models it reproduces the public Satorra-2000 difference test.

### Ordered and mixed indicators

Categorical support uses lavaan's inspected `UGamma` and unscaled estimator
statistic directly. DWLS (including WLSMV/WLSM/WLSMVS), ULS (including ULSMV),
and full WLS are supported experimentally for single- and multigroup ordered or
mixed-indicator models.

```{r categorical}
HSord <- HolzingerSwineford1939
ordered_names <- paste0("x", 1:9)
HSord[ordered_names] <- lapply(
  HSord[ordered_names],
  function(x) ordered(cut(x, 3))
)
fit_ordinal <- cfa(model, HSord, ordered = ordered_names)
pvalues(fit_ordinal, c("SB", "SS", "pEBA4"))
```

Categorical fits may use listwise or pairwise missingness. Pairwise support
means that semTests uses lavaan's pairwise sample-statistic and `UGamma`
calculation; it is not FIML and does not imply generally MAR-valid inference.
Nested categorical comparison supports `method = "2000"` with
`A.method = "delta"` only.

## References

Du, H., & Bentler, P. M. (2022). 40-Year Old Unbiased Distribution Free
Estimator Reliably Improves SEM Statistics for Nonnormal Data. *Structural
Equation Modeling*, 29(6), 872–887.
<https://doi.org/10.1080/10705511.2022.2063870>

Foldnes, N., Moss, J., & Grønneberg, S. (2025). Improved goodness of fit
procedures for structural equation models. *Structural Equation Modeling: A
Multidisciplinary Journal*, 32(1), 1–13.
<https://doi.org/10.1080/10705511.2024.2372028>

Foldnes, N., Grønneberg, S., & Moss, J. (2026). Penalized eigenvalue block
averaging: Extension to nested model comparison and Monte Carlo evaluations.
*Behavior Research Methods*, 58, article 107.
<https://doi.org/10.3758/s13428-026-02968-4>

Kenward, M. G., & Molenberghs, G. (1998). Likelihood based frequentist
inference when data are missing at random. *Statistical Science*, 13(3),
236–247. <https://doi.org/10.1214/ss/1028905886>

Satorra, A. (2000). Scaled and adjusted restricted tests in multi-sample
analysis of moment structures. In R. D. H. Heijmans, D. S. G. Pollock, &
A. Satorra (Eds.), *Innovations in Multivariate Statistical Analysis*
(pp. 233–247). Kluwer Academic.
<https://doi.org/10.1007/978-1-4615-4603-0_17>
