---
title: "Robust test statistics with semTests"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Robust test statistics with semTests}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
options(warn = -1)
```

## Introduction

`semTests` computes robust *p*-values for structural equation models fit with
[`lavaan`](https://lavaan.ugent.be/). The standard chi-square test of model fit
can reject too often under non-normality. Familiar corrections such as
Satorra--Bentler help. The eigenvalue-based tests in this package go directly
to the *limiting null law* of the test statistic. That law is a weighted sum of
independent chi-squares,
\(\sum_j \lambda_j \chi^2_1\), whose weights \(\lambda_j\) are the eigenvalues of
a matrix built from the model and the asymptotic covariance of the sample
moments. The same result holds for any minimum-discrepancy estimator. This is
why the package can cover normal-theory maximum likelihood, least squares,
categorical models, and FIML within one framework.

The penalized eigenvalue block-averaging (`pEBA`) and penalized regression
(`pOLS`) methods were introduced by Foldnes, Moss, and Grønneberg (2025).
Foldnes, Grønneberg, and Moss (2026) extended them to nested model comparison.

```{r setup}
library("semTests")
library("lavaan")
```

## Choose a focused guide

This vignette gives you the tour. Pick the focused guide that matches your
data when you want the full story:

* `vignette("continuous-data", package = "semTests")` covers complete
  continuous data, the classical ML workflow, GLS/ULS, and nested tests.
* `vignette("categorical-data", package = "semTests")` covers ordered and mixed
  indicators, pairwise missingness, and categorical nested comparison.
* `vignette("fiml-missing-data", package = "semTests")` covers continuous FIML,
  the finite-sample evidence for the observed-information default, the lavaan
  compatibility convention, and nested restriction maps.

## Basic usage: goodness of fit

Fit a model with `lavaan`, then call `pvalues()`. For ML models, request a
robust lavaan test such as `estimator = "MLM"` or `"MLR"`. This gives
`semTests` the asymptotic moment covariance used by the correction.

```{r basic}
model <- "visual  =~ x1 + x2 + x3
          textual =~ x4 + x5 + x6
          speed   =~ x7 + x8 + x9"
fit <- cfa(model, HolzingerSwineford1939, estimator = "MLM")
pvalues(fit)
```

The result is a named vector that also prints a one-line provenance footer
recording the estimator, data type, information matrix, and degrees of freedom
actually used. The full option record is available with
`attr(x, "semtests")`.

### The test-name grammar

Each requested test uses one of `TEST`, `TEST_UG`, `TEST_ML`, `TEST_RLS`,
`TEST_UG_ML`, or `TEST_UG_RLS`:

* **test** gives the family. Choose `pEBA<j>` for penalized block averaging,
  where `j` is usually between 2 and 6. Other choices include
  `pOLS<gamma>`, `pall`, `all`, `eba<j>`, `std`, `sb`, `ss`, and `sf`.
* **ug** asks for the unbiased Du--Bentler (2022) gamma. Without `UG`, the
  standard biased gamma is used.
* **ml / rls** chooses the base chi-square. `ML` uses the normal-theory
  discrepancy and `RLS` uses Browne's reweighted least squares statistic.
  Leave the suffix out to use the default for the fitted model.

Several tests can be requested at once:

```{r grammar}
pvalues(fit, c("SB_ML", "pEBA4_ML", "pOLS2_ML"))
```

The `UG` gamma and the `RLS` statistic belong to the classical case with
continuous, complete data and `estimator = "ML"`. Other model families get a
helpful error if either option is requested.

## Nested model comparison

To compare two nested models, fit both and pass them to `pvalues_nested()` with
the **constrained** model first. Here we test weak invariance of the textual
loadings:

```{r nested}
constrained <- "visual  =~ x1 + x2 + x3
                textual =~ a*x4 + a*x5 + a*x6
                speed   =~ x7 + x8 + x9"
m1 <- cfa(model, HolzingerSwineford1939, estimator = "MLM")
m0 <- cfa(constrained, HolzingerSwineford1939, estimator = "MLM")
pvalues_nested(m0, m1)
```

`pall` is the recommended default for nested comparison (Foldnes, Grønneberg,
and Moss, 2026). The reduction `method` is `"2000"` (Satorra 2000); the
Satorra--Bentler `"2001"` construction is temporarily unavailable in this
release. Both fits must use the same estimator, fitting conventions, variables,
groups, sample sizes, raw data, and missingness mask. Reversed inputs produce a
warning and are swapped.

## More data families

The same machinery reaches several other estimators. The non-ML paths below are
validated against an independent implementation and marked stable, alongside the
classical ML core. `?semTests-support` gives the complete list of supported
combinations.

GLS and ULS work directly (ULS needs a robust test for the asymptotic
covariance):

```{r estimators}
gls <- cfa(model, HolzingerSwineford1939, estimator = "GLS")
pvalues(gls, "pEBA4")

uls <- cfa(model, HolzingerSwineford1939, estimator = "ULS",
           test = "satorra.bentler")
pvalues(uls, "pEBA4")
```

Missing data is handled through full-information maximum likelihood. Fit with
`missing = "fiml"` and pass the fit to `pvalues()` as usual. FIML uses the
biased gamma and the standard statistic. The `UG` and `RLS` suffixes therefore
do not apply, and the suffix-free test names are the right choice.

```{r fiml}
HS <- HolzingerSwineford1939
set.seed(1)
HS$x1[sample(nrow(HS), 60)] <- NA
fit_fiml <- cfa(model, HS, missing = "fiml", estimator = "MLR")
pvalues(fit_fiml)
```

The default `fiml.convention = "observed"` uses observed saturated and model
information throughout, following Kenward and Molenberghs' (1998)
observed-information recommendation, and is independently validated against
magmaan. Use
`fiml.convention = "lavaan"` when exact reproduction of lavaan 0.7-2's inspected
single-model `UGamma` or nested Satorra-2000 correction is wanted:

```{r fiml-conventions}
pvalues(fit_fiml, fiml.convention = "observed")
pvalues(fit_fiml, fiml.convention = "lavaan")
```

For nested FIML comparisons, `A.method = "delta"` is the default and works
across a wider range of equivalent parameterizations. `"exact"` uses the
literal parameter restriction map and requires the same underlying full free
parameter set in both models.

For a single FIML model, the lavaan convention is
`lavInspect(fit, "UGamma")`. The scalar Yuan--Bentler--Mplus correction stored
by a default MLR fit is a different quantity and can differ. For nested FIML
models, the lavaan convention reproduces the public Satorra-2000 difference
test.

### Ordered and mixed indicators

Categorical support uses lavaan's inspected `UGamma` and unscaled estimator
statistic directly. DWLS (including WLSMV/WLSM/WLSMVS) and ULS (including ULSMV)
are supported for single- and multigroup ordered or
mixed-indicator models. Full WLS/ADF is refused, because its weight is already
the inverse moment covariance, so every correction collapses to the ordinary
chi-square.

```{r categorical}
HSord <- HolzingerSwineford1939
ordered_names <- paste0("x", 1:9)
HSord[ordered_names] <- lapply(
  HSord[ordered_names],
  function(x) ordered(cut(x, 3))
)
fit_ordinal <- cfa(model, HSord, ordered = ordered_names)
pvalues(fit_ordinal, c("SB", "SS", "pEBA4"))
```

Categorical fits may use listwise or pairwise missingness. Pairwise support
means that semTests uses lavaan's pairwise sample-statistic and `UGamma`
calculation. Pairwise deletion has its own missingness assumptions and does not
provide FIML or generally MAR-valid inference. Nested categorical comparison
supports `method = "2000"` with `A.method = "delta"` only.

## References

Du, H., & Bentler, P. M. (2022). 40-Year Old Unbiased Distribution Free
Estimator Reliably Improves SEM Statistics for Nonnormal Data. *Structural
Equation Modeling*, 29(6), 872–887.
<https://doi.org/10.1080/10705511.2022.2063870>

Foldnes, N., Moss, J., & Grønneberg, S. (2025). Improved goodness of fit
procedures for structural equation models. *Structural Equation Modeling: A
Multidisciplinary Journal*, 32(1), 1–13.
<https://doi.org/10.1080/10705511.2024.2372028>

Foldnes, N., Grønneberg, S., & Moss, J. (2026). Penalized eigenvalue block
averaging: Extension to nested model comparison and Monte Carlo evaluations.
*Behavior Research Methods*, 58, article 107.
<https://doi.org/10.3758/s13428-026-02968-4>

Kenward, M. G., & Molenberghs, G. (1998). Likelihood based frequentist
inference when data are missing at random. *Statistical Science*, 13(3),
236–247. <https://doi.org/10.1214/ss/1028905886>

Satorra, A. (2000). Scaled and adjusted restricted tests in multi-sample
analysis of moment structures. In R. D. H. Heijmans, D. S. G. Pollock, &
A. Satorra (Eds.), *Innovations in Multivariate Statistical Analysis*
(pp. 233–247). Kluwer Academic.
<https://doi.org/10.1007/978-1-4615-4603-0_17>
